---
title: "Identifying Expression Shifts from Ancestral Gene Expression Reconstructions"
author: "Erin Fry"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: cerulean
    highlight: espresso
    fig_width: 8
    fig_height: 6
    self_contained: false
---

**This script analyzes the Bayesian Ancestral Gene Expression Reconstructions (BAGERs) generated in BayesTraits at two ancestral nodes to identify expression shifts. In this case, I aim to identify expression shifts in the human lineage (since the common ancestor of Humans and Chimpanzees) in the human frontal cortex. The data analyzed in this document were published by [Brawand et al in 2011](http://www.nature.com/nature/journal/v478/n7369/abs/nature10532.html), collected from post-mortem frotal cortices of 6 primate species. This paper identified 4 genes with expression shifts in the human brain using maximum likelihood. The method that generated this data [can be found here](https://github.com/erinfry6/AncGeneExpReconstructionBT).**

**Genes with expression shifts were identified using the Bayesian Posterior Probability of Divergence (BPPD), which calculates what percentage of sampled iterations saw an increase or decrease in expression in the human lineage.**

## Setup
###Load paths and libraries
```{r set up paths and load libraries, warning=FALSE, message=FALSE}
## read in tissue argument

tissue<-"br"

## set paths to directories, be sure to modify your home directory and the Anccestral Reconstruction directory you are analyzing
path="/Users/lynchlab/Desktop/ErinFry/workflowr/AGER/"
pathData=paste(path,"data/forBAGER/",sep="")
pathResults=paste(path,"data/BAGERresults/",tissue,"/",sep="")
pathAncRecon=paste(pathResults,"AncStates/",sep="")

models<-c("VarRates","Lambda")

options(stringsAsFactors = FALSE)

library(biomartr)
library(dplyr)
library("ggplot2")
library(gplots)
library(plotly)
```

```{r Load AGERs list and gene information, echo=FALSE, cache=TRUE}
## define function for reading in these types of files which will read incorrectly using standard read.csv
read.tcsv = function(file, header=TRUE, sep=",", ...) {
  
  n = max(count.fields(file, sep=sep), na.rm=TRUE)
  x = readLines(file)
  
  .splitvar = function(x, sep, n) {
    var = unlist(strsplit(x, split=sep))
    length(var) = n
    return(var)
  }
  
  x = do.call(cbind, lapply(x, .splitvar, sep=sep, n=n))
  x = apply(x, 1, paste, collapse=sep) 
  out = read.csv(text=x, sep=sep, header=header, ...)
  return(out)
  
}


## set the nodes of interest
if (tissue=="br"){
  ancHominini="Node-00003"
  ancHomo="Node-00004"
  
} else if (tissue=="cb"){
  ancHominini="Node-00003"
  ancHomo="Node-00004"
  
} else if (tissue=="lv"){
  ancHominini="Node-00003"
  ancHomo="Node-00004"
  
} else if (tissue=="kd"){
  ancHominini="Node-00003"
  ancHomo="Node-00004"
  
} else if (tissue=="ts"){
  ancHominini="Node-00002"
  ancHomo="Node-00003"
  
} else if (tissue=="ht"){
  ancHominini="Node-00003"
  ancHomo="Node-00004"
  
} else {
  print("this tissue is not known") }


## set function for importing the ancestral expression reconstructions
read.AncRecon=function(file, firstrow, lastrow,header=F, sep='\t'){
  temp<-t(read.tcsv(file, sep='\t', header=F)[,firstrow:lastrow])
  colnames(temp)<-temp[1,]
  temp<-temp[-1,]
  return(temp)
}


prepforheatmap=function(matrix.or.df){
  dims <- dim(matrix.or.df)
matrix.or.df <- as.numeric(matrix.or.df)
dim(matrix.or.df) <- dims 
return(matrix.or.df)
}

```

### Write functions to visualize expression divergence

```{r set divergence and PPD plot functions}
## view the AGERs for each gene in a list of genes, lastit (last iteration) is the number of iterations sampled in the chain

viewPPDs<-function(genelist, m=Summary$modelchoice.model.choice,nbins=100){
  for (i in genelist){  #for each of these genes
  gene<- read.AncRecon(paste(pathAncRecon,m[i],"/",listcsv[i], sep=""), firstrow = it.begin, lastrow = (expectedrows+it.begin), sep='\t') # read reconstruction
    
  anc1<-as.numeric(gene[,which(colnames(gene)==paste(ancHomo, " - 1",sep=""))])
  anc2<-as.numeric(gene[,which(colnames(gene)==paste(ancHominini, " - 1",sep=""))])
  
  minimum=(min(anc1, anc2)) #minimum estimated value
    maximum=(max(anc1, anc2)) #maximum estimated
    breakby= (maximum-minimum)/nbins  #size of bins
    bins <- seq(minimum, maximum, by =breakby)  #set the bins for this gene
  hist(anc1,
       main= paste("Gene #", i,"-", Summary$hgnc_symbol[i],"- BPPD:",Summary$BayesianPostProbofDivergence[i]),  #title of the image
       xlab="Inferred Expression in sqrt(TPM)", 
       ylab="Number of Samples", 
       col="#1c9099", breaks=bins, xlim=c(minimum,maximum))
  hist(anc2, add=T, col="#737373", breaks=bins)
  hist(anc1, border="black", breaks=bins, add=T)
  #legend(300,300, c("Human-Chimpanzee", "Human"), col=c("#7fbf7b", "#af8dc3"), lwd=10, border=NULL)
  } }


## view the AGERs for each gene in a list of genes, lastit (last iteration) is the number of iterations sampled in the chain

viewDifference<-function(genelist,m=Summary$modelchoice.model.choice, lastit=1000,nbins=100){
  for (i in genelist){  #for each of these genes
  gene<- read.AncRecon(paste(pathAncRecon,m[i],"/",listcsv[i], sep=""), firstrow = it.begin, lastrow = (expectedrows+it.begin), sep='\t')# read reconstruction
  anc1<-as.numeric(gene[,which(colnames(gene)==paste(ancHomo, " - 1",sep=""))])
  anc2<-as.numeric(gene[,which(colnames(gene)==paste(ancHominini, " - 1",sep=""))])
  diffvector<-anc1-anc2
  hist(diffvector,
       main= paste("Gene #", i,"-", Summary$hgnc_symbol[i],"-Post Prob Divergence:",abs(max(1-Summary$BayesianPostProbofDivergence[i], Summary$BayesianPostProbofDivergence[i]))),  #title of the image
       xlab="AncHuman - AncHC (sqrtTPM)", 
       ylab="Number of Samples", col="#1c9099")
  } }



```

### Load and Examine the Results Summary file

```{r Load summary file, echo=FALSE, tidy=TRUE}

## uploaded saved file 
Summary<-read.table(paste(pathResults,"2017-07-18BAGERSummary.txt", sep=""),sep='\t', header=T)

head(Summary)

listcsv<-Summary$listcsv

## find which row the iteration information begins on for this tissue's tree
finding.information.about.file<-(read.tcsv(paste(pathAncRecon,models[length(models)],"/",listcsv[1],sep=""), sep='\t'))
it.begin<-which(colnames(finding.information.about.file)=="Itter")

## set the total number of rows you expect to have in each file so the code will warn you if reconstructions failed
expectedrows=ncol(finding.information.about.file)-it.begin

```

### Select genes that are expressed with successful reconstructions
Genes that succesfully reconstructed will have similar posterior variations of the two ancestral node reconstructions. Thus, to eliminate successfully reconstructed genes, you calculate the fold difference in standard deivation between the two reconstructions.

```{r reconstruction success, warnings=FALSE}
## the HC ancestral reconstruction generally has a higher standard deviation. If it is too much larger than the ancHuman the chain failed to reconstruct

## define divergence fold standard deviation cutoff
cutoffSD<-4

hist(log(Summary$foldSD), main="Reconstruction Fold Difference in Standard Deviation", xlab="log fold change in standard deviation between the AncHomini and AncHomo AGERS")
abline(v=log(cutoffSD), col="red")

## only select genes which are expressed, set cutoff for expression
## 2 TPM is a standard expression cutoff
expcut<-log(2)

## filter genes that did not converge or are not expressed in either reconstruction
reconsgenes<-filter(as.data.frame(Summary),foldSD<cutoffSD,MedianAncHominini>expcut | MedianAncHomo>expcut)

```

**`r (length(which(Summary$foldSD<cutoffSD))/nrow(Summary))*100`% of genes successfully reconstructed (with a cutoff of `r cutoffSD`x standard deviation fold change).**

**`r nrow(reconsgenes)` genes successfully reconstructed and are expressed.**


## Identify genes with expression shifts in the human lineage

**I have chosen to use the Bayesian Posterior Probability of Divergence to identify genes with different expression levels in ancHuman and ancHuman-Chimpanzee nodes. We will use a BPPD of 90% as a cutoff.**

```{r transcriptome qualities of converged reconstructions, echo=FALSE, warning=FALSE}
## define the BPPD cutoff to call a gene to have an expression shift
BPPDcutoff<-0.9

## identify all expressed, converged genes with expression shifts
shiftgenes<-filter(as.data.frame(Summary),BayesianPostProbofDivergence>BPPDcutoff,foldSD<cutoffSD,MedianAncHominini>expcut | MedianAncHomo>expcut)

write.table(shiftgenes,paste(pathResults,Sys.Date(),"90Cutoff.txt", sep=""),sep='\t')

## View the distribution of divergence of converged genes
hist(reconsgenes$BayesianPostProbofDivergence, #create histogram of percent divergence for tested genes
       main= "Transcriptome Divergence", #title
       xlab="Bayesian Posterior Probability of Divergence", 
       ylab="Number of Genes", cex.lab=1.3,
       col="#1c9099", breaks=100)
abline(v=BPPDcutoff, col="red")

## next look at scatter plot of reconstructions by the Posterior Probability of Divergence (the preferred measure of divergence)
# define divergence cutoffs
reconsgenes$BPPDLevel <- cut(reconsgenes$BayesianPostProbofDivergence, c(0,0.90,0.95,0.99,1.1))

BPPDcats<-reconsgenes %>% 
  group_by(BPPDLevel) %>%
  summarise(no_genes=length(BPPDLevel))
BPPDcats


# Reverse the levels and generate some labels for the legend
reconsgenes$labels <- factor(reconsgenes$BPPDLevel, levels = rev(levels(reconsgenes$BPPDLevel)),
                           labels = c('>99%', '95-99%',
                                       '90-95%','<90%'))
## split the data into layers
df_layer_1 <- reconsgenes[ reconsgenes$labels =="<90%",]
df_layer_90 <- reconsgenes[ reconsgenes$labels =="90-95%",]
df_layer_95 <- reconsgenes[ reconsgenes$labels =="95-99%",]
df_layer_99 <- reconsgenes[ reconsgenes$labels ==">99%",]


p<-ggplot(data=reconsgenes, mapping=aes(x=(MedianAncHomo),y=(MedianAncHominini), text=hgnc_symbol)) + 
  theme_bw() +
  theme(plot.title= element_text(size=15, face="bold"), axis.title.x=element_text(size=20),axis.title.y=element_text(size=20)) +
  ggtitle("Ancestral Transcript Levels and by BBPD in the Human Lineage") +
  labs( x="AncHominini sqrt(TPM)", y="AncHomo sqrt(TPM)", face="bold", size=20) +
  geom_point(data=df_layer_1, colour="#ffffcc", alpha=1) +
  geom_point(data=df_layer_90, colour="#a1dab4", alpha=1) +
  geom_point(data=df_layer_95, colour="#41b6c4", alpha=1) +
  geom_point(data=df_layer_99, colour="#0c2c84", alpha=1) +
  geom_abline(intercept=0,slope=1) + 
  scale_y_log10(limits=c(1e-1,1e2)) + scale_x_log10(limits=c(1e-1,1e2)) +
  geom_vline(xintercept=expcut, alpha=.5) + geom_hline(yintercept=expcut, alpha=0.5) 
p

ggplotly(p)
```

**`r nrow(shiftgenes)` were identified to have expression shifts in the human lineage with a Bayesian Posterior Probability of Divergence of 90% of greater.**


## Examine the ancestral gene expression reconstructions

To visiualize expression shifts, one can view the posterior probability divergences of the ancHuman (teal) and ancHC (gray).

### Of genes identified to have an expression shift in Brawand et al

```{r ID expression shifts and see how fits expectations}

## examine the divergence of genes identified by Brawand et al
for (g in c("LIX1", "SYT15","CENPT", "THBS4")){
viewPPDs(which(Summary$hgnc_symbol==g))
}

```


Both genes identified by Brawand et al in our dataset were identified using our method. 


### Of housekeeping genes

We do not expect expression shifts in housekeeping genes.

```{r}
## examine housekeeping gene divergence
## these are some housekeeping genes
housekeepingBPPD<-vector()
for (g in c("ATF1", "ATF2","ATF4", "BTF3","BAT1","HNRPK","EIF1","EIF1B","EIF2A","EIF4H","AARS","CARS","HARS","RPL8","RPS5","MRPL1","MRPL24","POLR1C","POLR1E","PFDN5","SNX5","H1FX","H2AFX")){
viewPPDs(which(Summary$hgnc_symbol==g))
  housekeepingBPPD[g]<-max(Summary$BayesianPostProbofDivergence[which(Summary$hgnc_symbol==g)], 0)
}

## what is the distribution of divergences of the housekeeping genes in this list
hist(housekeepingBPPD, breaks = 100, main="Distribution of Divergence of Housekeeping Genes", xlim=c(0,1))
abline(v=0.9)
```
**`r length(which(housekeepingBPPD>0.9))` Housekeeping genes of `r length(housekeepingBPPD)` tested were identified to have expression shifts, with a mean of `r mean(housekeepingBPPD)` BPPD.**


### Of newly identified genes with expression shifts using our method

```{r}
## view PPDs of genes identified with expression shifts
viewPPDs(which(Summary$hgnc_symbol %in% shiftgenes$hgnc_symbol)[1:10])
viewDifference(which(Summary$hgnc_symbol %in% shiftgenes$hgnc_symbol)[1:10])
``` 


## View original gene expression heatmaps of genes with expression shifts compared to one of all reconstructed genes

These heatmaps are generated using the expression data in all sampled primate frontal cortices. The first, of all reconstructed genes, does not elucidate any obvious patterns. The second, of all genes with expression shifts, shows clusters of genes with specific patterns. The most notable of which is either increased or decreased expression in humans compared to all other species.

```{r examine qualities of genes identified to have shifts in the original data, echo=FALSE}
## load expression data used to run scripts
#TPM_log<-t(read.table("TPM_log2.txt",header=F))
#colnames(TPM_log)<-TPM_log[1,]
#TPM_log<-TPM_log[-1,]
## to get it in the right format, save and reload
#write.table(TPM_log, "temp.txt")
TPM_log<-t(as.matrix(read.table(paste(pathData,tissue,"_exp.txt", sep=""),header=F)))

colnames(TPM_log)<-TPM_log[1,]
TPM_log<-TPM_log[-1,]

## select only converged genes to create heatmap
converged_expression<-prepforheatmap(TPM_log[(which(Summary$hgnc_symbol %in% reconsgenes$hgnc_symbol)-1),])

## load species color indicator
#categoricalData=c("hsa.br.F.1"="blue", "hsa.br.M.1"="blue", "hsa.br.M.2"="blue", "hsa.br.M.3"="blue", "hsa.br.M.4"="blue", "ptr.br.F.1"="#56B4E9","ptr.br.M.1"="#56B4E9", "ptr.br.M.2"="#56B4E9", "ptr.br.M.3"="#56B4E9", "ptr.br.M.4"="#56B4E9", "ptr.br.M.5"="#56B4E9","ppa.br.F.1"="#999999", "ppa.br.F.2"="#999999","ppa.br.M.1"="#999999","ggo.br.F.1"="#0072B2", "ggo.br.M.1"="#0072B2","ppy.br.F.1"="orange", "ppy.br.M.1"="orange","mml.br.M.2"="red")


heatmap.2(converged_expression, col=bluered(75),
          density.info="none", trace="none", symm=FALSE,
          dendrogram="row", scale="row",
          Rowv=T, Colv=F,colsep = 1:74, 
          sepcolor="black", 
          sepwidth=c(0.0000000001,0.000000001),
          labRow="", labCol="", main="Expression of Reconstructed Genes \n Human    Chimp    Bonobo    Gorilla   Orang   Mac")



## select only genes identified to have expression shifts

shift_expression<-prepforheatmap(TPM_log[(which(Summary$hgnc_symbol %in% shiftgenes$hgnc_symbol)-1),])


heatmap.2(shift_expression, col=bluered(75),
          density.info="none", trace="none", 
          dendrogram="row", scale="row",
          Rowv=T, Colv=F,colsep = 1:74, 
          sepcolor="black", 
          sepwidth=c(0.0000000001,0.000000001),
          labRow="", labCol="", main="Expression of all Genes with Expression Shifts \n Human    Chimp    Bonobo    Gorilla   Orang   Mac")

```


```{r ancrecon as mean exploration, warning=FALSE}
## human mean
human_mean<-apply(converged_expression[,grep(pattern = "hsa", colnames(TPM_log))], 1, mean)

## chimp mean
chimp_mean<-apply(converged_expression[,c(grep(pattern = "ptr", colnames(TPM_log)) , grep(pattern = "ppa", colnames(TPM_log)) )], 1, mean)

## human and chimp mean
HC_mean<-apply(converged_expression[,c(grep(pattern = "hsa", colnames(TPM_log)), grep(pattern = "ptr", colnames(TPM_log)) , grep(pattern = "ppa", colnames(TPM_log)) )], 1, mean)

non_human_mean<-apply(converged_expression[,-grep(pattern = "hsa", colnames(TPM_log))], 1, mean)

temp<-cbind(reconsgenes,human_mean, chimp_mean, HC_mean, non_human_mean)


df_layer_90 <- temp[ temp$labels =="90-95%",]
df_layer_95 <- temp[ temp$labels =="95-99%",]
df_layer_99 <- temp[ temp$labels ==">99%",]



## If BT is averaging the descendants to reconstuct the ancestor, we should see 1:1 between descendant mean and reconstructed mean

## of humans

ggplot(data=temp, mapping=aes(x=(MedianAncHomo),y=(human_mean), text=hgnc_symbol)) + 
  theme_bw() +
  theme(plot.title= element_text(size=15, face="bold"), axis.title.x=element_text(size=20),axis.title.y=element_text(size=20)) +
  ggtitle("Ancestral Transcript Levels and by BBPD in the Human Lineage") +
  labs( x="MedianAncHomo", y="human_mean", face="bold", size=20) +
  geom_point(colour="grey",alpha=0.5)+
    geom_point(data=df_layer_90, colour="#a1dab4", alpha=1) +
  geom_point(data=df_layer_95, colour="#41b6c4", alpha=1) +
  geom_point(data=df_layer_99, colour="#0c2c84", alpha=1) +
  scale_y_log10(limits=c(1e-1,1e2)) + scale_x_log10(limits=c(1e-1,1e2)) +
  geom_abline(intercept=0,slope=1) + 
  geom_vline(xintercept=expcut, alpha=1) + geom_hline(yintercept=expcut, alpha=1) 

## of humans and chimps

ggplot(data=temp, mapping=aes(x=(MedianAncHominini),y=(HC_mean), text=BayesianPostProbofDivergence)) + 
  theme_bw() +
  theme(plot.title= element_text(size=15, face="bold"), axis.title.x=element_text(size=20),axis.title.y=element_text(size=20)) +
  ggtitle("Ancestral Transcript Levels and by BBPD in the Human Lineage") +
  labs( x="MedianAncHominini", y="HC_mean", face="bold", size=20) +
  geom_point(colour="grey",alpha=0.5)+
    geom_point(data=df_layer_90, colour="#a1dab4", alpha=1) +
  geom_point(data=df_layer_95, colour="#41b6c4", alpha=1) +
  geom_point(data=df_layer_99, colour="#0c2c84", alpha=1) +
    scale_y_log10(limits=c(1e-1,1e2)) + scale_x_log10(limits=c(1e-1,1e2)) +
  geom_abline(intercept=0,slope=1) + 
  geom_vline(xintercept=expcut, alpha=1) + geom_hline(yintercept=expcut, alpha=1) 

## the reconstruction of humans is pretty much the mean. Of humans and chipms, there is some variation, but still many are the mean


## Another question: does the method ID genes with different gene expression in humans?

## look at only shift genes comparing human mean to mean of all non_humans

ggplot(data=temp, mapping=aes(x=(human_mean),y=(non_human_mean), text=hgnc_symbol)) + 
  theme_bw() +
  theme(plot.title= element_text(size=15, face="bold"), axis.title.x=element_text(size=20),axis.title.y=element_text(size=20)) +
  ggtitle("Ancestral Transcript Levels and by BBPD in the Human Lineage") +
  labs( x="human_mean", y="non_human_mean", face="bold", size=20) +
  geom_point(colour="grey",alpha=0.5)+
    geom_point(data=df_layer_90, colour="#a1dab4", alpha=1) +
  geom_point(data=df_layer_95, colour="#41b6c4", alpha=1) +
  geom_point(data=df_layer_99, colour="#0c2c84", alpha=1) +
    scale_y_log10(limits=c(1e-1,1e2)) + scale_x_log10(limits=c(1e-1,1e2)) +
  geom_abline(intercept=0,slope=1) + 
  geom_vline(xintercept=expcut, alpha=1) + geom_hline(yintercept=expcut, alpha=1) 

## it is identifying genes with differing means in the original dataset, and giving us a way to model the gene's evolution


## if the reconstructions are basically equal to the mean, we should see the humand and HC mean's diverging for genes with shifts

ggplot(data=temp, mapping=aes(x=(human_mean),y=(HC_mean), text=hgnc_symbol)) + 
  theme_bw() +
  theme(plot.title= element_text(size=15, face="bold"), axis.title.x=element_text(size=20),axis.title.y=element_text(size=20)) +
  ggtitle("Ancestral Transcript Levels and by BBPD in the Human Lineage") +
  labs( x="human_mean", y="HC_mean", face="bold", size=20) +
  geom_point(colour="grey",alpha=0.5)+
    geom_point(data=df_layer_90, colour="#a1dab4", alpha=1) +
  geom_point(data=df_layer_95, colour="#41b6c4", alpha=1) +
  geom_point(data=df_layer_99, colour="#0c2c84", alpha=1) +
    scale_y_log10(limits=c(1e-3,1e2)) + scale_x_log10(limits=c(1e-3,1e2)) +
  geom_abline(intercept=0,slope=1) + 
  geom_vline(xintercept=expcut, alpha=1) + geom_hline(yintercept=expcut, alpha=1) 

##this signal coming from the chimpanzees

ggplot(data=temp, mapping=aes(x=(human_mean),y=(chimp_mean), text=hgnc_symbol)) + 
  theme_bw() +
  theme(plot.title= element_text(size=15, face="bold"), axis.title.x=element_text(size=20),axis.title.y=element_text(size=20)) +
  ggtitle("Ancestral Transcript Levels and by BBPD in the Human Lineage") +
  labs( x="human_mean", y="chimp_mean", face="bold", size=20) +
  geom_point(colour="grey",alpha=0.5)+
    geom_point(data=df_layer_90, colour="#a1dab4", alpha=1) +
  geom_point(data=df_layer_95, colour="#41b6c4", alpha=1) +
  geom_point(data=df_layer_99, colour="#0c2c84", alpha=1) +
    scale_y_log10(limits=c(1e-3,1e2)) + scale_x_log10(limits=c(1e-3,1e2)) +
  geom_abline(intercept=0,slope=1) + 
  geom_vline(xintercept=expcut, alpha=1) + geom_hline(yintercept=expcut, alpha=1) 

```
